---
title: "CT5100 Assignment 4 Part 2"
author: "Conor O'Donoghue"
date: "April 12, 2021"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Explanation
  To get the x-axis labels correct I created a new factor, country, which is just the LOCATION factor translated to country names using countrycode.
I then wanted to order the country factor by the boys' pisa values as in the plot.
this was trickier than anticipated -- I found that I could use dplyr's fct_reorder2() to
reorder the country factor by both subject (boy/girl) and value, but found that whether I included or excluded a - sign it 
ended up sorting by the girls' PISA scores. The fix was to turn subject into a numeric first, THEN trying to sort by subject descending
and value ascending worked.

  I wanted to use the same colors as the original plot, so I obtained the hex values for the boys' points, the outline of the girls' points,
the background, the red, the black, and the grey that the rest of the text was colored in using imagecolorpicker.com. I couldn't find a programmatic way to
change the text colors of the x labels, so I made a vector of grey values and adjusted the values at the indexes where the Irish and OECD values would be
to red and black respectively.

  I plotted the vertical lines for each country first and the points after, so that the vertical lines wouldn't draw through the circles or the diamonds.
The white lines used geom_segment() to draw a line from the lower limit (340) to each point, and geom_line() to draw a line connecting the boys' and girls' values for each country.
I highlighted the OECD Average values in black and the Irish values in red by avoiding plotting them with the first geom_point() call,
and then using additional geom_point() calls to explicitly plot each set of points with the right colors.

  I changed the breaks in scale_y_continuous to start at 340, end at 560, and increment by 20, and set the limits to be just outside that.
I removed most aspects of the theme using element_blank(). Notably, I left in the horozontal gridlines by setting panel.grid.major.y equal to element_line()
with a white color, set panel.background to an element_rect() with the same color as the background color of the original plot, 
rotated the x-axis labels and set the colors with the angle= and color= arguments in element_text() for axis.text.x,
and moved the y-axis labels up with axis.text.y = element_text(vjust=-0.3).

  I had to set the legend's position manually and nudge the plot's subtitle and caption where I wanted them
using hjust and vjust, and reduced the size of all the text until it was roughly proportional
to the original plot.

  The one aspect of the plot that I was unable to reproduce was the background (including the light blue color and the gridlines)
extending left into the y-axis labels. It should have to do with the theme() function, but I couldn't figure out a way to make it work.
The best I could do was use vjust() to move the y-axis labels up so that they would be just above each gridline if I could have extended it left.

\pagebreak
# Plot
```{r, echo=FALSE, message=FALSE}
# Loading libraries
library(ggplot2)
library(dplyr)
library(forcats)
library(countrycode)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Reading in data
df <- read.csv('./OECD_PISA.csv')

# Subsetting to 2018 mean scores only
df_18 <- df[which(df$TIME == 2018 & df$SUBJECT %in% c("BOY", "GIRL")),]

# Adding country names
df_18$country <- countrycode(df_18$LOCATION, origin='iso3c', destination='country.name')
df_18[df_18$LOCATION=="OAVG", "country"] <- "OECD - Average"

# Order Location by BOY'S value
df_18 <- df_18 %>% mutate(country = fct_reorder2(country, -as.numeric(SUBJECT), Value, .desc=FALSE))

# Colors (obtained with imagecolorpicker.com)
bg_color = "#DEEDF3"
black_color = "#000000"
red_color = "#FF0024"
boy_color = "#326A88"
girl_color = "#3C6F8D"

# Set up text color
# for the life of me I could not find a programmatic way to do the text color.
# I was trying to use the location factor LEVELS to make a new array of colors with "grey" for all but "red" where IRL was and "black" where OAVG was,
# But I couldn't figure it out. So I hard coded it.
text_color = rep("#9b9797", 41)
text_color[21] <- black_color
text_color[40] <- red_color

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Recreating plot
ggplot(df_18, aes(x = reorder(country,Value), y = Value)) +
        coord_fixed(ratio=0.07) +
        # plot white lines from bottom of the plot to each point
        geom_segment(aes(x=country, y=340, xend=country, yend=Value), color="white", size=0.5) +
        # plot grey lines between the boys and the girls points for each country
        geom_line(aes(group = country), colour = "grey", size=0.5) +
        # plot averages as black
        geom_point(aes(shape=SUBJECT), fill=black_color, color=black_color, size=2, data=df_18[df_18$LOCATION == "OAVG",]) +
        # plot ireland as red
        geom_point(aes(shape=SUBJECT), fill=red_color, color=red_color, size=2, data=df_18[df_18$LOCATION == "IRL",]) +
        # plot the remainder of the points
        geom_point(aes(shape = SUBJECT, fill= SUBJECT, color=SUBJECT), size=2, data=df_18[!df_18$LOCATION %in% c("OAVG", "IRL"),]) +
        # Rather than 'no fill', can also fill girl color with same color as the background.
        scale_shape_manual(values=c(21, 23), labels=c("Boys", "Girls")) +
        scale_color_manual(values=c(boy_color,girl_color), labels=c("Boys", "Girls")) +
        scale_fill_manual(values= c(boy_color,bg_color), labels=c("Boys", "Girls")) +
        # Start and end must be 20 units higher than breaks
        scale_y_continuous(limits = c(320, 580), 
                           expand = c(0, 0),
                           minor_breaks=seq(340,560,by=20),
                           breaks = seq(340, 560, by = 20),
                           name = "seconds") +
        labs(title="Reading performance (PISA)",
             subtitle="Boys/Girls, Mean score, 2018",
             caption="Source: PISA: Programme for International Student Assessment") +
              # Move title, subtitle, and caption to match original plot
        theme(plot.title = element_text(size=10, vjust=-5),
              plot.subtitle = element_text(size=6.5, hjust=0.38, vjust=1),
              plot.caption.position = "plot",
              plot.caption = element_text(size=6.5, vjust=166.5),
              panel.grid.major.x = element_blank(),
              panel.grid.minor.y = element_blank(),
              # Draw horozontal gridlines
              panel.grid.major.y = element_line(color="white", size=0.5),
              # Set background color
              panel.background = element_rect(fill=bg_color),
              # Remove axis lines and ticks
              axis.line.y = element_blank(),
              axis.line.x = element_blank(),
              axis.ticks.y = element_blank(),
              axis.ticks.x = element_blank(),
              # Add x labels, rotate, color accordingly
              axis.text.x = element_text(size=6, angle=45, hjust=1, color=text_color),
              axis.title.x = element_blank(),
              axis.text.y = element_text(size=6, vjust=-0.3),
              axis.title.y = element_blank(),
              # set legend position to bottom left, set entries to be horizontal
              legend.position= c(0.05, -0.25),
              legend.direction = "horizontal",
              legend.text = element_text(size = 7),
              legend.key.size = unit(0.7, "lines"),
              legend.background = element_blank(),
              legend.title = element_blank())
```


\pagebreak
# Code Appendix
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```